---
title: "11. Linear model extensions"
author: "jsg"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
---

```{r}
require(tidyverse)
require(lme4)
require(car)
require(multcomp)
require(Rmisc)
library(MuMIn)
require(mgcv)
```

Before doing this, review the ** Linear Model Extensions** lecture set slides 
from https://sites.google.com/view/biostats/lessons/linear-model-extensions
and the  **11_Mixed_models_and_other_extensions.R**
script in the lecture files folder of the
[CUNY-BioStats github repository](https://github.com/jsgosnell/CUNY-BioStats). 
Make sure you are comfortable with null and alternative hypotheses and appropriate plots
for all examples.

Remember you should

* add code chunks by clicking the *Insert Chunk* button on the toolbar or by
pressing *Ctrl+Alt+I* to answer the questions!
* **knit** your file to produce a markdown version that you can see!
* save your work often 
  * **commit** it via git!
  * **push** updates to github

1. In a study considering how the presence of sea stars changed snail growth 
patterns, ~25 snails were grown in containers containing 0,1, or 2 seastars.  
Since non-consumptive effects are often threshold based, these treatments levels
should be considered as groups (not as a continuous variable!).  The data is 
available at

https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/snail_modified_for_class.csv   

FL is the final length of measured snails, and the treatment (coded 1-3) correspond
to  [1=Control (no predators). 2=1 predator treatment,3=2 predator treatment). 

What method would you use to analyze this data and why? Carry out your test, 
stating your null hypothesis, test assumptions, p-value, and interpretation.  
Describe any necessary steps and provide graphics and values as needed.  If 
needed, can you determine which treatments differ from each other?

```{r}
snail <- read.csv(file = "snail_modified_for_class.csv") %>%
as_tibble() %>%
  mutate_at(vars(Treatment), factor) %>%
    mutate(treat = case_when(
    Treatment == "1" ~ "Control", 
    Treatment == "2" ~ "Single predator",
    Treatment == "3" ~ "Two predators"
  )
    ) %>%
    mutate_at(vars(treat), factor)
```

```{r}
snail
```

```{r}
snail.lmer <- lmer(FL ~ treat + (1|Container), snail)
```

```{r}
summary(snail.lmer)
```

```{r}
plot(snail.lmer)
```

```{r}
check_mixed_model <- function (model, model_name = NULL) {
  #collection of things you might check for mixed model
  par(mfrow = c(2,3))
  #not sure what this does with mutliple random effects, so stop with 1 for now
  if(length(names(ranef(model))<2)){
    qqnorm(ranef(model, drop = T)[[1]], pch = 19, las = 1, cex = 1.4, main= paste(model_name, 
                                                                                  "\n Random effects Q-Q plot"))
  }
  plot(fitted(model),residuals(model), main = paste(model_name, 
                                                    "\n residuals vs fitted"))
  qqnorm(residuals(model), main =paste(model_name, 
                                       "\nresiduals q-q plot"))
  qqline(residuals(model))
  hist(residuals(model), main = paste(model_name, 
                                      "\nresidual histogram"))
}
```

```{r}
check_mixed_model(snail.lmer)
```

```{r}
Anova(snail.lmer, type = "III")
```

```{r}
snail_comparison <- glht(snail.lmer, linfct = mcp(treat = "Tukey"))
```

```{r}
summary(snail_comparison)
```

```{r}
#graph using Rmisc
snail_summary <- summarySE(snail, measurevar = "FL", groupvars = "treat")
```

```{r}
snail_summary
```

```{r}
snail.barplot <- ggplot(snail_summary, aes(x = treat, y = FL)) +   
  geom_bar(stat = "identity", width = 0.75, color = "#2b2b2b", linewidth = 0.50, alpha = 0.6) +
  geom_linerange(aes(ymin = FL - ci, ymax = FL + ci), linewidth = 0.75) +
  #ylab(expression(paste("Gas (1000 ",ft^3, ")")))+
  xlab("Treatment") +
  #geom_smooth(method = "lm", se = F) +
#  scale_y_continuous(expression(paste("Temperature")), limits = c(0, 10)) + 
#  scale_x_discrete(expand = c(0, 1)
                   #, 
                   #labels = x_labels
#                   ) + 
#  scale_fill_manual(breaks = c("North", "West",
#                               "Southeast"),
#                    values = c("red", "blue", 
#                               "green"), 
#                    labels = c("North", "West",
#                               "Southeast")) +
#  facet_wrap( ~ Insulate, labeller = as_labeller(label_names), dir = "v", ncol = 1) + 
  ggtitle(expression(paste(italic(" Response Varibale: FL")))) +
  #scale_y_log10(expression(paste("Colony Size (", cm^2, ")"), limits = c(0, 100000))) +
  labs(x = NULL) +
  theme(#strip.text = element_text(size = 10, color = "black", hjust = 0.50),
        #strip.background = element_rect(fill = "#FFFFFF", color = NA),    
        #panel.background = element_rect(fill = "#FFFFFF", color = NA),
        #panel.grid.major.x = element_blank(),
        #panel.grid.minor.x = element_blank(),
        #panel.grid.minor.y = element_blank(),
        #panel.grid.major.y = element_line(color = "#b2b2b2"),
        #panel.spacing.x = unit(1, "cm"),
        #panel.spacing.y = unit(0.5, "cm"),
        #panel.spacing = unit(1, "lines"),
        axis.ticks = element_blank(),
        #legend.position = "top",
        plot.title = element_text(size = 11),
        axis.title.y = element_text(size = 11),
        legend.title = element_blank()
        )
```

```{r}
snail.barplot
```

**Conclusion**

*Since multiple oysters were measured in each cage, we need to use a random effect
to account for cages. You could also block by cages- it takes up more degrees of 
freedom, but you have plenty here.  Results show a significant differce among 
treatments (Chi^2~2=10.681, p <.01), so I used a Tukey post hoc test to determine 
which groups differed from others while controlling for the family wise error rate.
REsults indicate the presence of a predator impacts length but not the density.*


2. (From OZDasl) The data give the ambient temperature and the number of 
primary O-rings damaged for 23 of the 24 space shuttle launches before the 
launch of the space shuttle Challenger on January 20, 1986. (Challenger was the
25th shuttle. One engine was lost at sea and could not be examined.) Each space
shuttle contains 6 primary O-rings.

Note these are counts. We can analyze this data using a Poisson distribution 
or binomial. Make sure you understand why each one is possible, which one is 
better, and carry out the analysis.  Data is available @ 

http://www.statsci.org/data/general/challenger.txt

```{r}
rings <- read.table("http://www.statsci.org/data/general/challenger.txt", 
                    header = T)
```

```{r}
rings
```

```{r}
challenger <- read.csv(file = "challenger.csv") %>%
as_tibble() 
#%>%
#  mutate_at(vars(Treatment), factor) %>%
#    mutate(treat = case_when(
#    Treatment == "1" ~ "Control", 
#    Treatment == "2" ~ "Single predator",
#    Treatment == "3" ~ "Two predators"
#  )
#    ) %>%
#    mutate_at(vars(treat), factor)
```

```{r}
challenger
```

```{r}
# Poisson with whole numbers, i.e., counts
challenger.glm1 <- glm(Damaged ~ Temp, family = "poisson", data = challenger)
```

```{r}
par(mfrow = c(2, 2))
plot(challenger.glm1)
```

```{r}
summary(challenger.glm1)
```

**Note**: Dispersion is ok!!

```{r}
Anova(challenger.glm1, type = "III")
```

*Try Binomial -> proportion of successes:failures <- number damaged and not!*

```{r}
challenger2 <-
  challenger %>%
  mutate(
    No_damage = 6 - Damaged)
```

```{r}
challenger2
```

```{r}
challenger.glm2 <- glm(cbind(Damaged, No_damage) ~ Temp, family = "binomial", data = challenger2)
```

```{r}
par(mfrow = c(2, 2))
plot(challenger.glm2)
```

```{r}
Anova(challenger.glm2, type = "III")
```

**Compare to** *lm*

```{r}
challenger.lm <- lm(Damaged ~ Temp, data = challenger2)
```

```{r}
par(mfrow = c(2, 2))
plot(challenger.lm)
```

```{r}
summary(challenger.lm)
```

**Note**: Dispersion is ok!!

```{r}
Anova(challenger.lm, type = "III")
```

*Since these are counts we need to use a glm to model the data. We could use a 
Poisson, but the binomial actually includes more information (like how many did
not fail!). Both models indicate a significant relationship between temperature
and the number or proportion of failed rings. Results are compared to a linear model.*

3. Returning to the whelk length-mass relationship from class, try fitting an 
exponential curve to the data.  As a hint, try

```{r eval=F}
nls(Mass ~ exp(b0 + b1 * Shell.Length), whelk, 
                   start = list(b0 =1, b1=0), na.action = na.omit)
```

Compare this model to those that assume a linear and power relationship.  Data is available @

https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/whelk.csv 

```{r}
whelk <- read.csv(file = "whelk.csv") %>%
as_tibble() %>%
  mutate_at(vars(Location, Sex), factor) %>%
  mutate_at(vars(Mass, Shell.Length), as.numeric)
#    mutate(treat = case_when(
#    Treatment == "1" ~ "Control", 
#    Treatment == "2" ~ "Single predator",
#    Treatment == "3" ~ "Two predators"
#  )
#    ) %>%
#    mutate_at(vars(treat), factor)
```

```{r}
whelk
```

```{r}
whelk.reg <- ggplot(data = whelk, aes(x = Shell.Length, y = Mass, color = Location)) +   
  geom_point(size = 2, alpha = 0.6) +
#  ylim (0.0, 1.0) +
  ggtitle(expression(paste(italic(" Whelk")))) +
#  xlab(expression(paste("Mean Wave Height (m)"))) + 
#  ylab(expression(paste("Benthic Coverage"))) +
#  annotate(
#    "text",
#    x = Inf, y = -Inf,
#    label = eqn_plob, parse = TRUE,
#    hjust = 1.1, vjust = -.5
#  ) +
  theme_bw() + 
#  facet_wrap( ~ type,  ncol = 3) +
  theme(#panel.spacing = unit(1, "lines"),
        legend.position = "right",
        plot.title = element_text(size = 11),
        legend.title = element_blank()) 
#+
 # stat_smooth(method = "lm", se = FALSE, color = "black")
```

```{r}
whelk.reg
```

```{r}
whelk_plot <- ggplot(whelk, aes_string(x="Shell.Length", y = "Mass")) +
  geom_point(aes_string(colour = "Location")) + 
  theme(axis.title.x = element_text(face="bold", size=28), 
        axis.title.y = element_text(face="bold", size=28), 
        axis.text.y  = element_text(size=20),
        axis.text.x  = element_text(size=20), 
        legend.text =element_text(size=20),
        legend.title = element_text(size=20, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size=32))
```

```{r}
whelk_plot
```

*Power fit*

```{r}
whelk.lm <- lm(Mass ~ Shell.Length, whelk, na.action = na.omit)
```

```{r}
whelk.power <- nls(Mass ~ b0 * Shell.Length^b1, whelk, 
                   start = list(b0 = 1, b1 = 3), na.action = na.omit)
```

```{r}
whelk.exponential <- nls(Mass ~ exp(b0 + b1 * Shell.Length), whelk, 
                         start = list(b0 = 1, b1 = 0), na.action = na.omit)
```

```{r}
AICc(whelk.lm, whelk.power, whelk.exponential)
```

**Now, plot!**

```{r}
whelk_plot + 
  
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.5, color = "orange") + 
  
  geom_smooth(method = "nls", 
              # look at whelk_power$call
              formula = y ~ b0 * x^b1, 
              method.args = list(start = list(b0 = 1, 
                                              b1 = 3)), 
              se = FALSE, linewidth = 1.5, color = "blue") +
  geom_smooth(method = "nls", 
              # look at whelk_exponential$call
              formula = y ~ exp(b0 + b1 * x), 
              method.args = list(start = list(b0 = 1, 
                                              b1 = 0)), 
              se = FALSE, linewidth = 1.5, color = "green")
```

*We can use the nls model to consider exponential curve to the data. Various fits
may be compared using AIC methods.  In this case it appears that the power fit is 
the best (lowest AIC value).*

4. Going back to the TEAM dataset, remember we found that elevation had no 
impact on carbon storage.  But that was a linear fit. Use a gam (generalized 
additive model) to see if elevation can be related to carbon storage in an 
additive model.  Note we can use the gamm (generalized additive mixed model) 
function in the mgcv package to denote mixed effects.  For example (from help 
file)

```{r eval = F}
b2 <- gamm(y ~ s(x0) + s(x1) + s(x2) , family = poisson,
           data = dat, random = list(fac = ~1))
```

Team data is available @ 

https://github.com/jsgosnell/CUNY-BioStats/blob/master/datasets/team_data_no_spaces.csv 

```{r}
team <- read.csv("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/team_data_no_spaces.csv", stringsAsFactors = T)
```

```{r}
team
```

```{r}
elevation_linear <- gam(PlotCarbon.tonnes ~ Elevation, data = team)
```

```{r}
elevation_gam <- gam(PlotCarbon.tonnes ~ s(Elevation), data = team)
```

```{r}
elevation_gamm <- gamm(PlotCarbon.tonnes ~s(Elevation), random = list(Site.Name = ~ 1), data = team)
```

```{r}
AICc(elevation_gam, elevation_gamm, elevation_linear)
```




