---
title: "6. Continuous tests 2 populations"
author: "jjh"
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y %H:%M')`"
output:
  html_document:
    toc: true
    toc_float: true
    keep_md: true
    self_contained: true
    toc_depth: 6
---

```{r}
library(reshape)
library(reshape2)
library(coin)
library(tidyverse)
packageVersion("dplyr")
require(Rmisc)
require(BSDA)
require(simpleboot)
```

## Practice

### For the following problems, use the data to conduct each noted test (note some sample sizes may be too small for these to all be good ideas!)


#### 1

1. The following data are human blood clotting times (in minutes) of individuals
given one of two different drugs. 

|Drug B | Drug G
|:-: |:-:
|8.8 | 9.9
|8.4 | 9.0
|7.9 | 11.1
|8.7 | 9.6
|9.1 | 8.7
|9.6 | 10.4 
|    | 9.5

Test the hypothesis that the mean clotting times
are equal for the two groups

* Estimating the variance from the data 
* Using rank transform analysis 
* Using a permutation test 
* Using a bootstrap test 

```{r}
drug.df <- data.frame(
  
  drugB = c(8.8, 8.4, 7.9, 8.7, 9.1, 9.6, NA),
  
  drugG = c(9.9, 9.0, 11.1, 9.6, 8.7, 10.4, 9.5)
  
)
```

```{r}
drug3.df <-
  drug.df %>%
  pivot_longer(
    cols = drugB:drugG,
    names_to = "type",
    values_to = "drug"
  ) %>%
  mutate(type = as.factor(type))
```

```{r}
drug3.df
```

```{r}
summarySE(drug3.df, measurevar = 'drug', groupvars = 'type', na.rm = T)
```

```{r}
wilcox.test(drug ~ type, drug3.df, na.action = "na.pass")
```

```{r}
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/bootstrapjsg.R")
```

```{r}
bootstrapjsg(drug.df$drugB, drug.df$DrugG)
```

```{r}
independence_test(drug ~ type, drug3.df)
```

#### 2

2. Data on plant heights (in cm) for plants grown with a new and old formulation 
of fertilizer can be found at
https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/fertilizer.csv . 
Use the data to test the hypothesis that there is no difference in mean plant
heights for the two groups 

* Estimating the variance from the data 
* Using rank transform analysis 
* Using a permutation test 
* Using a bootstrap test 

Note you may need to cast this dataframe into a dataframe with a column for 
old and new outcomes.  You’ll have to add a unique identifier. Try something like

```{r, eval=FALSE}
fertilizer$ID <- 1:nrow(fertilizer)
fertilizer_wide <- dcast(fertilizer, ID~fertilizer, value.var = "height")
```

```{r}
fert <- read.csv('fertilizer.csv')
```

```{r}
fert
```

```{r}
fert$ID <- 1:nrow(fert)
fertilizer_wide <- dcast(fert, ID ~ fertilizer, value.var = "height")
```

```{r}
summarySE(fert, measurevar = 'height', groupvars = 'fertilizer', na.rm = T)
```

```{r}
wilcox.test(drug ~ type, drug3.df, na.action = "na.pass")
```

```{r}
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/bootstrapjsg.R")
```

```{r}
bootstrapjsg(fertilizer_wide$new, fertilizer_wide$old)
```

```{r}
independence_test(drug ~ type, drug3.df)
```

#### 3

3. Metabolic rates were measured in animals before and after administering a drug. Data is
available at 

https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/metabolic_rates.csv

Test the hypothesis that the drug has no effect on metabolic rate with using 

* a Student's t test
* a rank test
* a binary test
* bootstrapping

```{r}
meta <- read.csv('meta.csv')
```

```{r}
meta
```

#### A Student's t test

```{r}
t.test(meta$before, meta$after, paired = T)
```

#### A rank test

```{r}
wilcox.test(meta$before, meta$after, paired = T)
```

```{r}
meta_diff <- na.omit(meta$before - meta$after)
bootstrapjsg(meta_diff)
```

#### 4

4. Endurance of the above animals was also measured. Data is available at

https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/datasets/endurance.csv

Test the hypothesis that the drug has no effect on using 
* a Student's t test
* a rank test
* a binary test
* bootstrapping

```{r}
meta <- read.csv('meta.csv')
```

```{r}
meta
```

#### A Student's t test

```{r}
t.test(meta$before, meta$after, paired = T)
```

#### A rank test

```{r}
wilcox.test(meta$before, meta$after, paired = T)
```

#### A Binary test



#### Bootstrapping

```{r}
meta_diff <- na.omit(meta$before - meta$after)
bootstrapjsg(meta_diff)
```


### For the following problems, make sure you can justify the test you choose to use and carry it out

#### 5

5. Using data from Australian athletes (http://www.statsci.org/data/oz/ais.html 
for details), determine if the average male training at the Australian Institute 
of Sport differs in from the average female training at the site in mercury
concentration (Hg). 

```{r}
ais <- read.csv('ais.csv')
```

```{r}
ais
```

```{r}
ais_tibble <-
  as_tibble(ais) %>%
  mutate_at(c('Sex', 'Sport'), as.factor)
```

```{r}
ais_tibble
```

```{r}
ais_histogram <- 
  ais_tibble %>%
  ggplot(aes(Hg)) +
  geom_histogram() # +
  #facet_wrap(vars(Sex), ncol = 1)
```

```{r}
ais_histogram
```

```{r}
ais_histogram <- 
  ais_tibble %>%
  ggplot(aes(Hg)) +
  geom_histogram() +
  facet_wrap(vars(Sex), ncol = 1)
```

```{r}
ais_histogram
```

```{r}
summarySE(ais_tibble, measurevar = 'Hg', groupvars = c('Sex', 'Sport'), na.rm = T)
```

```{r}
summarySE(ais_tibble, measurevar = 'Hg', groupvars = c('Sex'), na.rm = T)
```

We can assume equal variances between **males** and **females**.

##### T-test

```{r}
t.test(Hg ~ Sex, ais_tibble)
```

##### Rank test

```{r}
wilcox.test(Hg ~ Sex, ais_tibble)
```

**Binary test**

```{r}
# SIGN.TEST()
```

```{r}
ais_wide <-
  ais_tibble %>%
  select(Sex, Hg) %>%
  group_by(Sex) %>%
  dplyr::mutate(row = row_number()) %>%
  tidyr::pivot_wider(names_from = Sex, 
              values_from = Hg)
```

```{r}
ais_wide
```

##### Bootstrapping

```{r}
source("https://raw.githubusercontent.com/jsgosnell/CUNY-BioStats/master/code_examples/bootstrapjsg.R")
```

#```{r}
#bootstrapjsg(ais_wide$female, ais_wide$male)
#```

##### Permutation

```{r}
independence_test(Hg ~ Sex, 
                  ais_tibble)
```

#### 6

6. Laureysens et al. (2004) measured metal content in the wood of 13 poplar 
clones growing in a polluted area, once in August and once in November.
Concentrations of aluminum (in micrograms of Al per gram of wood) are shown 
below. (From Handbook of Biological Statistics). Has metal content changed over
time?

| Clone | August | November
|:-:| :-:|:-:
|Columbia River | 18.3 |12.7
|Fritiz Pauley | 13.3 |11.1
|Hazendas | 16.5 |15.3
|Primo | 12.6 | 12.7
|Raspalje | 9.5 |10.5
|Hoogvorst |13.6 |15.6
|Balsam Spire | 8.1 |11.2
|Gibecq   |8.9 |14.2
|Beaupre | 10.0 |16.3
|Unal |8.3 |15.5
|Trichobel |7.9 | 19.9
|Gaver | 8.1 | 20.4
|Wolterson | 13.4 |36.8

```{r}
clone.df <- data.frame(
  
  Clone = c("Columbia River", "Fritiz Pauley", "Hazendas", "Primo", "Raspalje",
            "Hoogvorst", "Balsam Spire", "Gibecq", "Baupre", "Unal", "Trichobel",
            "Gaver", "Wolterson"),
  
  August = c(18.3, 13.3, 16.5, 12.6, 9.5, 13.6, 8.1, 8.9, 10.0, 8.3, 7.9, 8.1, 13.4),
  
  November = c(12.7, 11.1, 15.3, 12.7, 10.5, 15.6, 11.2, 14.2, 16.3, 15.5, 19.9, 20.4, 36.8)
  
)
```

```{r}
clone.df
```

```{r}
clone_long <-
  clone.df %>%
  pivot_longer(!Clone, names_to = "Date", values_to = "Concentration"
  )
```

```{r}
clone_long
```

```{r}
ggplot(clone_long, aes(x = Date, y = Concentration)) +
  geom_point(color = "purple", size = 2) +
  ggtitle("Metal content in the wood") +
  theme(axis.title.x = element_text(face="bold", size = 14), 
        axis.title.y = element_text(face="bold", size = 14), 
        axis.text.y  = element_text(size = 10),
        axis.text.x  = element_text(size = 10), 
        legend.text =element_text(size = 10),
        legend.title = element_text(size = 10, face = "bold"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 18))
```

```{r}
clone_difference <-
  clone.df %>%
    mutate(diff = November - August)
```

```{r}
clone_difference
```

```{r}
ggplot(clone_difference, aes(x = "diff", y = diff)) +
  geom_point(color = "purple", size = 2) +
  ggtitle("Metal content in the wood") +
  theme(axis.title.x = element_text(face = "bold", size = 14), 
        axis.title.y = element_text(face = "bold", size = 14), 
        axis.text.y  = element_text(size = 10),
        axis.text.x  = element_text(size = 10), 
        legend.text =element_text(size = 10),
        legend.title = element_text(size = 10, face = "bold"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 18))
```

```{r}
#confidence interval####
difference_summary <- summarySE(clone_difference, measurevar = "diff")
```

```{r}
difference_summary
```

```{r}
ggplot(difference_summary, aes(x = .id, y = diff)) +
  geom_point(color = "purple", size = 2) +
  geom_errorbar(aes(ymin = diff - ci, ymax = diff + ci))+
  ggtitle("Metal content in the wood") +
  theme(axis.title.x = element_text(face = "bold", size = 14), 
        axis.title.y = element_text(face="bold", size = 14), 
        axis.text.y  = element_text(size = 10),
        axis.text.x  = element_text(size = 10), 
        legend.text =element_text(size = 10),
        legend.title = element_text(size = 10, face="bold"),
        plot.title = element_text(hjust = 0.5, face="bold", size = 18)) +
  xlab(NULL)+
  ylab("Difference in metal content")
```

```{r}
#example of paired impacts
t.test(clone.df$August, clone.df$November)
```

```{r}
#or
t.test(Concentration ~ Date, clone_long)
```

```{r}
#vs paired
t.test(clone.df$August, clone.df$November, paired = T)
```

```{r}
#or
t.test(Concentration ~ Date, clone_long, paired = T) #assumes in proper order!
```

#### 7

7. Data was collected on cholesterol levels following heart attacks.  Data
available at 
http://www.statsci.org/data/general/cholestg.txt 

Note to read a .txt file you use,
read.table("http://www.statsci.org/data/general/cholestg.txt", header = T, 
stringsAsFactors = T)
Check the defaults to understand the argument.  
More info on the data is @
http://www.statsci.org/data/general/cholest.html
Is there evidence cholesterol levels changed between day 2 and day 4?  
* **THINK ABOUT RELABELLING DAY AS A FACTOR (WHY?)?**

```{r}
chol <- read.csv('cholestg.csv')
chol
```

```{r}
chol_tibble <-
  as_tibble(chol) %>%
  mutate_at(c('group', 'day'), as.factor) %>%
  mutate_at(c('patient'), as.character)

chol_tibble
```

```{r}
chol_wide <-
  chol_tibble %>%
  pivot_wider(names_from = "day", values_from = "cholest") %>%
  dplyr::rename("day2" = "2", "day4" = "4") %>%
  select(-'14', -"NA") %>%
  na.omit()

chol_wide
```

```{r}
chol_long <-
  chol_wide %>%
  select(-'patient') %>%
  pivot_longer(!group, names_to = "day", values_to = "chol")

chol_long
```

```{r}
ggplot(chol_long, aes(x = day, y = chol)) +
  geom_point(color = "purple", size = 2) +
  ggtitle("Days and Cholesterol") +
  theme(axis.title.x = element_text(face="bold", size = 12), 
        axis.title.y = element_text(face="bold", size = 12), 
        axis.text.y  = element_text(size = 10),
        axis.text.x  = element_text(size = 10), 
        legend.text =element_text(size = 10),
        legend.title = element_text(size = 10, face = "bold"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

```{r}
chol_difference <-
  chol_wide %>%
  mutate(diff = day4 - day2)
chol_difference
```

```{r}
ggplot(chol_difference, aes(x = "diff", y = diff)) +
  geom_point(color = "purple", size = 2) +
  ggtitle("Days and cholesterol") +
  theme(axis.title.x = element_text(face = "bold", size = 12), 
        axis.title.y = element_text(face = "bold", size = 12), 
        axis.text.y  = element_text(size = 10),
        axis.text.x  = element_text(size = 10), 
        legend.text =element_text(size = 10),
        legend.title = element_text(size = 10, face = "bold"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14))
```

```{r}
chol_dif_summary <- summarySE(chol_difference, measurevar = "diff")
chol_dif_summary
```

```{r}
ggplot(chol_dif_summary, aes(x = .id, y = diff)) +
  geom_point(color = "purple", size = 2) +
  geom_errorbar(aes(ymin = diff - ci, ymax = diff + ci))+
  ggtitle("Difference: Days and cholesterol") +
  theme(axis.title.x = element_text(face = "bold", size = 12), 
        axis.title.y = element_text(face="bold", size = 12), 
        axis.text.y  = element_text(size = 10),
        axis.text.x  = element_text(size = 10), 
        legend.text =element_text(size = 10),
        legend.title = element_text(size = 10, face = "bold"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14)) +
  xlab(NULL)+
  ylab("Difference in cholesterol")

```

```{r}
# Pivot Wider
t.test(chol_wide$day2, chol_wide$day4)
```

```{r}
# Pivot Longer
t.test(chol ~ day, chol_long)
```

```{r}
# Paired pivot wider

t.test(chol_wide$day2, chol_wide$day4, paired = T)
```

```{r}
#or

t.test(chol ~ day, chol_long, paired = T) #assumes in proper order!
```


